### 2019-09-16 ic-journal/expansion.tex
Removed all reviewed highlights in the first part of the paper; re-enable highlights

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \hl{(weak)} \emph{bisimulation} if, whenever
 $P\RR Q$\hl{,} % for all $\mu$ we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\hl{;}
\end{enumerate}
 $P$ and $Q$ are \hl{(weakly)} \emph{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

\hl{\emph{Strong} bisimulation and strong bisimilarity} ($\sim$)
can be obtained by replacing the weak transition $Q\Arcap\mu Q'$ in the above
definition with normal transitions $Q \arr \mu Q'$ (similar for the other case).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums), i.e.\  $P_1 \wb Q_1$ and $P_2 \wb Q_2$  \hl{does not imply}
 $P_1 + P_2 \wb Q_1 + Q_2$.
For this, Milner introduced the concept of \emph{observational congruence}, also called \emph{rooted
  bisimilarity}~\cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, \hl{if} % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

\hl{The above definition also brings up a proof technique for proving rooted
bisimilarity from a given bisimulation. The next
lemma plays an
important role in proving some key results in this paper:}
\begin{lemma}[Rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
\hl{Given a (weak) bisimulation $\RR$, suppose two processes $P$ and $Q$
satisfy the following properties}:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ \hl{implies that there is} $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
\hl{Then} $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

One basic property of rooted bisimiarity is congruence,
indeed the coarsest congruence contained in bisimilarity
(c.f. Section~\ref{s:coarsest} for more details):
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

>>>>

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a (weak) \emph{bisimulation} if, whenever
 $P\RR Q$, % for all $\mu$ we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$;
\end{enumerate}
 $P$ and $Q$ are (weakly) \emph{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

\emph{Strong} bisimulation and strong bisimilarity ($\sim$)
can be obtained by replacing the weak transition $Q\Arcap\mu Q'$ in the above
definition with normal transitions $Q \arr \mu Q'$ (similar for the other case).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums), i.e.\  $P_1 \wb Q_1$ and $P_2 \wb Q_2$  \hl{do} not imply
 $P_1 + P_2 \wb Q_1 + Q_2$.
For this, Milner introduced the concept of \emph{observational congruence}, also called \emph{rooted
  bisimilarity}~\cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, if % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

The above definition also brings up a proof technique for proving rooted
bisimilarity from a given bisimulation. The next
lemma plays an
important role in proving some key results in this paper:
\begin{lemma}[Rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, suppose two processes $P$ and $Q$
satisfy the following properties:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
Then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

One basic property of rooted bisimiarity is congruence,
indeed the coarsest congruence contained in bisimilarity
(c.f. Section~\ref{s:coarsest} and \hl{\mbox{Section~\ref{ss:context}}} for more details):
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

### 2019-09-16 ic-journal/intro.tex
Removed all reviewed highlights in the first part of the paper; re-enable highlights

\hl{Milner's unique-solution theorem for weak bisimilarity, however,
has severe syntactic limitations:}
the equations must be both \emph{guarded} and \emph{sequential}. That is,
variables of the equations \hl{can only occur underneath visible prefixes and summations.}
One way \hl{to overcome} such limitations is to replace the equations
with special inequations called
\emph{contractions}~\cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency constraints on processes.
\hl{The unique solution of contractions is defined as with equations:}
any two solutions must be componentwise bisimilar.
The difference \hl{from} equations is \hl{in} the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem \hl{are
eliminated}. One constraint that still remains
in~\cite{sangiorgi2017equations}
(where the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of %\emph{direct} 
\hl{sums} (e.g.~$P + Q$) due to
the \hl{non-substitutivity of the contraction preorder in this case}.

 \hl{This paper presents}  a
comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
\hl{Many results in Milner's CCS book}~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong and weak
\hl{bisimilarities} (e.g.~the fixed-point and substitutivity properties), the
basic properties of rooted bisimilarity and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, the expansion and contraction preorder.
Concerning rooted bisimilarity, the formalisation
includes Hennessy's Lemma and Deng's Lemma,
 and two theorems saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other without such hypothesis, essentially formalised van Glabbeek's proof~\cite{van2005characterisation}.
Similar results are also proved for the rooted contraction preorder.
In this respect, the work is also an extensive experiment using the HOL theorem prover with its
recent developments, including its coinduction defining package.

From the point of view of the CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular, the existing result~\cite{sangiorgi2017equations} has limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
In this paper, the contraction-based proof technique is further
refined by moving from the contraction preorder to
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
\hl{constraints on the use the sum operator}.

@@

We first consider the case of a single equation (or contraction),
that we call the \univariate case, then consider the \multivariate
  case, where  more than one equation
  (or contraction) \hl{with multiple equation variables is considered.}
The \univariate theorems use
$\lambda$-functions to present CCS equations,
while the \multivariate versions require a
more careful and delicate treatment of CCS
expressions with multiple variables and substitutions acting on them.

>>>>

Milner's unique-solution theorem for weak bisimilarity, however,
has severe syntactic limitations:
the equations must be both \emph{guarded} and \emph{sequential}. That is,
variables of the equations can only occur underneath visible prefixes and summations.
One way to overcome such limitations is to replace the equations
with special inequations called
\emph{contractions}~\cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency constraints on processes.
The unique solution of contractions is defined as with equations:
any two solutions must be componentwise bisimilar.
The difference from equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem are
eliminated. One constraint that still remains
in~\cite{sangiorgi2017equations}
(where the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of sums (e.g.~$P + Q$) due to
the non-substitutivity of the contraction preorder in this case.

This paper presents a comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
Many results in Milner's CCS book~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong and weak
bisimilarities (e.g.~the fixed-point and substitutivity properties), the
basic properties of rooted bisimilarity and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, the expansion and contraction preorder.
Concerning rooted bisimilarity, the formalisation
includes Hennessy's Lemma and Deng's Lemma,
 and two theorems saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other without such hypothesis, essentially formalised van Glabbeek's proof~\cite{van2005characterisation}.
Similar results are also proved for the rooted contraction preorder.
In this respect, the work is also an extensive experiment using the HOL theorem prover with its
recent developments, including its coinduction defining package.

From the point of view of the CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular, the existing result~\cite{sangiorgi2017equations} has limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
In this paper, the contraction-based proof technique is further
refined by moving from the contraction preorder to
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the use the sum operator.

@@

We first consider the case of a single equation (or contraction),
that we call the \univariate case, then consider the \multivariate
  case, where  more than one equation
  (or contraction) with multiple equation variables is considered.
The \univariate theorems use
$\lambda$-functions to present CCS equations,
while the \multivariate versions require a
more careful and delicate treatment of CCS
expressions with multiple variables and substitutions acting on them.
